---
title: "Lab 3"
author: "Clint Valentine"
output: pdf_document
---
  
Modeling the Dynamics of Structured Populations
-----------------------------------------------
  
We will use the Lotka-Volterra model for direct competition between two species as described below in the system of differential equations describing $N_{1}$ and $N_{2}$:
  
\begin{center}1.1\hfill$\frac{dN_{1}}{dt} = r_{1}N_{1}(\frac{K_{1}-N_{1}-\alpha_{12}N_{2}}{K_{1}})$\hfill\phantom{.}\end{center}

\begin{center}1.2\hfill$\frac{dN_{2}}{dt} = r_{1}N_{2}(\frac{K_{2}-N_{2}-\alpha_{21}N_{1}}{K_{2}})$\hfill\phantom{.}\end{center}

This system of differential equations is easily solvable using the `ode45` algorithm. An example system is proposed, solved, and plotted.

````{r fig.width=7, fig.height=5}
library(deSolve)
library(manipulate)
library(fields)

source('http://faraway.neu.edu/data/cb.R')

solve.lotka.volterra <- function(t, y, parms) {
  dX <- parms$r1 * y[1] * ((parms$K1 - y[1] - parms$alpha12 * y[2]))
  dY <- parms$r2 * y[2] * ((parms$K2 - y[2] - parms$alpha21 * y[1]))
  return(list(c(dX, dY)))
}


N = c(32, 32)
parms = list(K1=200, K2=110, r1=2, r2=2, alpha12=2, alpha21=2)
times <- 1:10

results <- ode(y=N, times=times, func=solve.lotka.volterra, parms,
               method="ode45")

matplot(results[,2:3], type='l', lty=1, xlab='Time', ylab='Abundance',
        col=c(cb.red, cb.blue))

legend(x='right', legend=c(expression('N'[1]), expression('N'[2])),
       lty=c(1,1), lwd=c(2.5,2.5), col=c(cb.red, cb.blue))
```

To graphically display the phase diagram for a dual species Lotka-Volterra model we will first define the zero net growth isoclines of the system. We will then simulate many starting points of the system and draw their solutions as lines on the phase diagram.

````{r fig.width=7, fig.height=5}
plot.competition.ZNGI <- function(grid, time.length,
                                  K1, K2, r1, r2, alpha12, alpha21) {
  
  parms <- list(K1=K1, K2=K2, r1=r1, r2=r2, alpha12=alpha12, alpha21=alpha21)
  times <- seq(0, 1, len=time.length)
  
  N1 <- matrix(
    c(0, (parms$K1 / parms$alpha12), parms$K1, 0), nrow=2, ncol=2, byrow=TRUE)
  N2 <- matrix(
    c(0, parms$K2, (parms$K2 / parms$alpha21), 0), nrow=2, ncol=2, byrow=TRUE)
  
  xmax <- max(parms$K1, parms$K2 / parms$alpha21) + 1
  ymax <- max(parms$K2, parms$K1 / parms$alpha12) + 1
  
  x.equil <- (parms$K1 - parms$alpha12 * parms$K2) /
    (1 - parms$alpha12 * parms$alpha21)
  
  y.equil <- (parms$K2 - parms$alpha21 * parms$K1) /
    (1 - parms$alpha12 * parms$alpha21)
  
  N1.vect = seq(0.001, max(parms$K1, parms$K2 / parms$alpha21), len=grid)
  N2.vect = seq(0.001, max(parms$K2, parms$K1 / parms$alpha12), len=grid)
  
  initials <- expand.grid(N1.vect, N2.vect)
  
  plot(NULL, xlim=c(0,xmax), ylim=c(0,ymax), xlab=expression('N'[1]),
       ylab=expression('N'[2]), main='Zero Net Growth Isoclines for Two Species')
  
  for (x in 1:length(initials[,1])) {
    N <- c(initials[x, 1], initials[x, 2])
    results <- ode(y=N, times=times, func=solve.lotka.volterra, parms,
                   method='ode45')
    lines(results[,2], results[,3], type='l', lty=1)
  }
  
  lines(N1, type='l', lty=1, lwd=3, col=cb.blue)
  lines(N2, type='l', lty=1, lwd=3, col=cb.red)
  
  if (N1[1, 2] < N2[1, 2]) {
    y.boundary <- N2[1, 2]
    points(0, y.boundary, col=cb.yellow, pch=16)
    points(x.equil, y.equil, col=cb.yellow, pch='o')
  } else if (N1[1, 2] > N2[1, 2]) {
    y.boundary <- N1[1, 2]
    points(0, y.boundary, col=cb.yellow, pch='o')
    points(x.equil, y.equil, col=cb.yellow, pch=16)
  }
  
  if (N1[2, 1] < N2[2, 1]) {
    x.boundary <- N2[2, 1]
    points(x.boundary, 0, col=cb.yellow, pch='o')
  } else if (N1[2, 1] > N2[2, 1]) {
    x.boundary <- N1[2, 1]
    points(x.boundary, 0, col=cb.yellow, pch=16)
  }
  
  legend(x='topright', legend=c(expression('N'[1]), expression('N'[2])),
         lty=c(1,1), lwd=c(2.5,2.5), col=c(cb.blue, cb.red))
}

plot.competition.ZNGI(grid=20, time.length=100, 120, 40, 0.5, 0.35, 2, 3)
```
The following code was used to manipulate the zero net growth isocline plotting function in real time:
  ````{r fig.width=7, fig.height=5, eval=FALSE}
manipulate ({
  plot.competition.ZNGI(gridVal, Tlen, K1Val,
                        K2Val, r1Val, r2Val, alpha12Val, alpha21Val)
},
gridVal=slider(min=5, max=100, initial=20, label="Grid Dimension"),
Tlen=slider(min=100, max=10000, initial=100, label="Timestep Resolution"),
K1Val=slider(min=20, max=500, initial=100, label="K1 Value"),
K2Val=slider(min=20, max=500, initial=100, label="K2 Value"),
r1Val=slider(min=0, max=4, initial=0.35, step=0.05, label="Rate for N1"),
r2Val=slider(min=0, max=4, initial=0.35, step=0.05, label="Rate for N2"),
alpha12Val=slider(min=0, max=30, initial=2, label="Alpha12"),
alpha21Val=slider(min=0, max=30, initial=2, label="Alpha21")
)
```

Stability Analysis
------------------
  
  An eigen analysis of the Jacobian matrix of the five potential equilibrium points will also elucidate the behaviour of the Lotka-Volterra interaction. First, the three boundary conditions are tested. The extinction condition at $N_{1} = 0$ and $N_{2} = 0$ is stable only if $r_{1}$ and $r_{2}$ are greater than zero. Either monoculture boundary condition are stable if the species growth rate is positive and if the ratio of the competitor's carrying capacity to the species' carrying capacity is smaller than the competitive effect of the species on the competitor.

The interior equilbirium is only present if it exists in the domain of biological relevance (positive integers). This point is stable if the intraspecific competition of both species is greater than the interspecific species.

If both species have stable monoculture points than their are two alternative stable equilibriums in the system with a potential unstable coexistence point.

The following functions determine the characteristics of a system graphically and numerically using eigen analysis of the Jacobian matrix.
````{r fig.width=7, fig.height=5}

make.Jacob <- function(N1, N2, K1, K2, r1, r2, alpha12, alpha21) {
  J <- matrix(
    data=c(
      (r1 / K1) * (K1 - (2 * N1) - (alpha12 * N2)),
      (-r1 / K1) * (alpha12 * N1),
      (-r2 / K2) * (alpha21 * N2),
      (r2 / K2) * (K2 - (2 * N2) - (alpha21 * N1))),
    nrow=2, ncol=2, byrow=FALSE)
  return(J)
}

test.Stability <- function(K1, K2, r1, r2, alpha12, alpha21) {
  
  b.origin <- Re(eigen(make.Jacob(0,  0, K1, K2, r1, r2, alpha12, alpha21))$values)
  
  if (b.origin[1] < 0 && b.origin[2] < 0)  {
    return(1)
  }
  
  if (alpha12 * alpha21 != 1) {
    N1.equil <- (K1 - alpha12 * K2) / (1 - alpha12 * alpha21)
    N2.equil <- (K2 - alpha21 * K1) / (1 - alpha12 * alpha21)
    if (N1.equil > 0 && N2.equil > 0) {
      i.equil <- Re(eigen(make.Jacob(N1.equil, N2.equil, K1, K2, r1, r2,
                                     alpha12, alpha21))$values)
      if (i.equil[1] < 0 && i.equil[2] < 0) {
        return(4)
      }
    }
  }
  
  b.N1 <- Re(eigen(make.Jacob(K1, 0, K1, K2, r1, r2, alpha12, alpha21))$values)
  b.N2 <- Re(eigen(make.Jacob(0, K2, K1, K2, r1, r2, alpha12, alpha21))$values)
  
  N1.mono <- (b.N1[1] < 0 & b.N1[2] < 0)
  N2.mono <- (b.N2[1] < 0 & b.N2[2] < 0)
  
  if (N1.mono & N2.mono) {
    return(5)
  } else if (N1.mono) {
    return(2)
  } else if (N2.mono) {
    return(3)
  }
}

test.Stability(50, 23, 3, 4, 3, 1) # 5 = Multiple Stable States
plot.competition.ZNGI(20, 1000, 50, 23, 3, 4, 3, 1)
```

We will now simulate this system for all variations of the following parameter values: $\alpha_{12} = 0.5$, $\alpha_{21}$ ranging from $0$ to $2.5$, $K_{1} = 3$, $K_{2}$ ranging from $0.01$ to $15$, $r_{1} = r_{2} = 2$. These results are then stored into a matrix for future plotting.

````{r fig.width=7, fig.height=5}
source("http://faraway.neu.edu/data/format3d.R")

p <- expand.grid(alpha12=0.5, alpha21=seq(0, 2.5, len=100), r1=2, r2=2,
                 K1=3, K2=seq(0.01, 15, len=100))

results = matrix(nrow = nrow(p), ncol=3)

for (x in 1:nrow(p)) {
  pr <- p[x,]
  results[x,] <- cbind(pr$K2/pr$K1,
                       pr$alpha21/pr$alpha12,
                       test.Stability(pr$K1, pr$K2, pr$r1, pr$r2,
                                      pr$alpha12, pr$alpha21))
}

results.3d <- format3d(results)
image(results.3d, xlab=expression(frac(K[2], K[1])), main='Stability Analysis of Parameter Space', col=tim.colors(4))

text(x=-0.25, y=2.5, expression(frac(alpha[21], alpha[12])), xpd=TRUE, srt=0, pos=2)
text(0.75, 3.7, col='white', expression(paste(N[1], ' monoculture')))
text(1.25, 0.8, col='white', 'Alternative States')
text(2.15,  4.8, col='white', 'Co.')
text(3.5,  2.5, col='white', expression(paste(N[2], ' monoculture')))

```

This plot makes sense. I ran a few simulations with paramter values in this space and I saw the corresponding zero net growth isoclines indicitave of these scenarios. When $K_{1}$ is larger than the value of $K_{1}$ there are two outcomes determined by the ratio of the species' respective interspecific competitive effect. We see an $N_{1}$ monoculture when $\alpha_{21} > \alpha_{12}$ and alternative monocultures of the two species when $\alpha_{21} < \alpha_{12}$ depending on the initial population values.

When $K_{2}$ is larger than the value of $K_{1}$ there are also two outcomes that are determined by the ratio of the species' respective interspecific competitive effect. When $\alpha_{21} > \alpha_{12}$ we see coexistance of the two species and when $\alpha_{21} < \alpha_{12}$ we see a complete $N_{2}$ monoculture.


Extending and Simulating Mechanistic Models of Competition
----------------------------------------------------------

To extend the model above to mechanisitic competition by modeling the common resource between the two species we must use a system of three differential equations. Instead of using parameters that simulate a direct competitive effect on each species we will look at indirect competition through a common essential resource.

###These are not right at all
\begin{center}1.1\hfill$\frac{dN_{1}}{dt} = r_{1}N_{1}(\frac{K_{1}-N_{1}-\alpha_{12}N_{2}}{K_{1}})$\hfill\phantom{.}\end{center}

\begin{center}1.2\hfill$\frac{dN_{2}}{dt} = r_{1}N_{2}(\frac{K_{2}-N_{2}-\alpha_{21}N_{1}}{K_{2}})$\hfill\phantom{.}\end{center}

````{r fig.width=7, fig.height=5}
library(deSolve)

solve.mechanistic <- function(t, y, parms) {
  dR <- parms$r1 * (parms$S1 - y[1]) - (y[2] * parms$u1 * y[1]) / (parms$Y1 * (parms$K1 + y[1])) - (y[3] * parms$u2 * y[1]) / (parms$Y2 * (parms$K2 + y[1]))
  dX <- y[2] * (parms$u1 * (y[1] / (parms$K1 + y[1])) - parms$M1)
  dY <- y[3] * (parms$u2 * (y[1] / (parms$K2 + y[1])) - parms$M2)
  return(list(c(dR, dX, dY)))
}

N = c(100, 20, 30)
parms = list(r1=4, S1=100, K1=32, u1=3, M1=2, K2=32, u2=5, M2=3.4, Y1=5, Y2=9)
times <- 1:25

results <- ode(y=N, times=times, func=solve.mechanistic, parms,
               method="ode45")

matplot(results[,2:4], type='l', lty=1, xlab='Time', ylab='Abundance',
        col=c(cb.green, cb.blue, cb.red))

legend(x='topleft', legend=c('R', expression('N'[1]), expression('N'[2])),
       lty=c(1,1,1), lwd=c(2.5,2.5,2.5), col=c(cb.green, cb.blue, cb.red))


R1.critical <- (parms$K1 * parms$M1) / (parms$u1 - parms$M1)
R2.critical <- (parms$K2 * parms$M2) / (parms$u2 - parms$M2)

print(R1.critical)
print(R2.critical)



```
