---
title: "Assignment 12A"
author: "Clint Valentine"
date: "April 16, 2017"
output: pdf_document
---

```{r, message=FALSE, warning=FALSE}
rm(list = ls())

library(jsonlite)
library(mongolite)
library(stringr)
```


```{r}
# Load in the data into one data.frame if it has not already been loaded.
if (exists('bird.strikes') == F) {
  unzip('Bird Strikes.zip')
  # Escape sep = ',' in quotations
  # Impute NA as empty strings
  # Do not consider strings factors
  bird.strikes <- read.table(
    file = 'Bird Strikes.csv',
    fill = T,
    header = T,
    na.strings = '',
    quote = '\"',
    sep = ',',
    stringsAsFactors = F
  )
}

# Replace periods with underscore, remove trailing underscore, and make lowercase.
header <- gsub('_$', '', gsub('\\.+', '_', str_trim(colnames(bird.strikes))))
header <- tolower(header)

# Replace reformatted header on bird.strike data.
colnames(bird.strikes) <- header
```


Insert the Bird Strikes.csv file into MongoDB
---

```{r}
mongo <- mongo(collection='test',
               db='test',
               url='mongodb://localhost:27017',
               verbose=F)

mongo$insert(bird.strikes)
```

\newpage

Use the export command to display the inserted file.
---

1. Data is first exported to a tempfile.
2. Since this ouputs many valid JSON lines but not in a valid JSON array we must preprocess the information.
3. Read into a `data.frame`.
4. Examine results by calling `dim` and `str`.

Note there is an extra column in our exported dataset which represents the unique identifer MongoDB has assigned our record.

```{r, message=FALSE, warning=FALSE}
tmp <- tempfile()
mongo$export(file(tmp))
exported.data <- fromJSON(sprintf("[%s]", paste(readLines(file(tmp)), collapse=",")))
```
```{r}
print(dim(bird.strikes))
print(dim(exported.data))
```


```{r}
str(exported.data, width=80, strict.width='cut')
```

Fetch the unique airport names from the database
---

```{r}
airports <- mongo$distinct('airport_name')
cat("There are", length(airports), "unique airport names.\n\n")
cat("Here are the first ten:\n\n")
print(head(airports, 10))
```

Count the number of records where `origin_state` equals "New Jersey".
---
```{r}
query <- '{
  "origin_state": "New Jersey"
}'

num_new_jersey <- mongo$count(query)
cat("There are", num_new_jersey, "records where `origin_state` = New Jersey.")
```

\newpage

Fetch the data with `conditions_precipitation` being fog and sort the data in descending order of `record_id`.
---

```{r}
query <- '{
  "conditions_precipitation": "Fog"
}'

sort <- '{
  "record_id": -1
}'

records <- mongo$find(query=query, sort=sort)
cat("There are", nrow(records), "records with `conditions_precipitation` as Fog.\n\n")
cat("Are the records sorted in descending order of `record_id`?",
    all.equal(records$record_id,
              rev(sort(records$record_id))))
```

\newpage

Fetch only the following columns for `aircraft_airline_operator`
---

Query:

`aircraft_airline_operator` = "AMERICAN AIRLINES" OR "CONTINENTAL AIRLINES"

Columns to include:

- `record_id`
- `origin_state`
- `aircraft_airline_operator`
- `airport_name`

```{r}
query = '{
  "aircraft_airline_operator": { "$in": ["AMERICAN AIRLINES", "CONTINENTAL AIRLINES"]}
}'

fields = '{
  "_id": 0,
  "record_id": 1,
  "origin_state": 1,
  "aircraft_airline_operator": 1,
  "airport_name": 1
}'

records <- mongo$find(query=query, fields=fields)
cat("There are", nrow(records), "records for `AMERICAN AIRLINES`/`CONTINENTAL AIRLINES`.")
print(head(records))
```
